\documentclass[a4paper,12pt]{extarticle}

% =========================
% Page layout (Annex 7)
% =========================
\usepackage[
  left=2.5cm,
  right=1.0cm,
  top=2.0cm,
  bottom=2.0cm
]{geometry}

% =========================
% Fonts + languages
% =========================
\usepackage{fontspec}
\setmainfont{CMU Serif}
\setsansfont{CMU Sans Serif}
\setmonofont{CMU Typewriter Text}

\usepackage{polyglossia}
\setdefaultlanguage{english}
\setotherlanguage{russian}

\newfontfamily\russianfont{CMU Serif}
\newfontfamily\russianfontsf{CMU Sans Serif}
\newfontfamily\russianfonttt{CMU Typewriter Text}

% =========================
% Math + text
% =========================
\usepackage{listings}
\usepackage{booktabs}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{csquotes}
\usepackage{graphicx}
\usepackage{indentfirst}
\usepackage{setspace}
\usepackage{microtype}

% =========================
% Figures / plots
% =========================
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

\usepackage{caption}
\DeclareCaptionLabelSeparator{emdash}{\space\textemdash\space}
\captionsetup{
  font=small,
  labelfont=bf,
  labelsep=emdash,
  justification=centering
}

\graphicspath{{graphics/}}

% =========================
% Bibliography
% =========================
\usepackage[
  backend=bibtex,
  style=numeric,
  maxbibnames=99,
  urldate=long
]{biblatex}
\addbibresource{refs.bib}

% =========================
% hyperref — loaded ONCE with all options
% =========================
\usepackage[
  colorlinks=true,
  citecolor=blue,
  linkcolor=blue,
  urlcolor=blue,
  bookmarks=false
]{hyperref}

% =========================
% Formatting
% =========================
\setlength{\parindent}{1.25cm}
\onehalfspacing
\pagestyle{plain}
\numberwithin{equation}{section}

\begin{document}

% 1. Title page
\input{title_kr}
\clearpage

\setcounter{page}{2}

% 2. Contents
{
  \hypersetup{linkcolor=black}
  \tableofcontents
}
\clearpage

% 3. Abstract
\section*{Abstract}
\addcontentsline{toc}{section}{Abstract}
This project presents a complete black-box A/B testing pipeline applied to
a UI button experiment. The joint system of UI element and user is treated
as a black box: the input is the assignment of a user to control variant A or
treatment variant B, and the outputs are product metrics (click-through rate,
CTR) computed from logged user events. The work delivers a formalised
experimental methodology (unit of randomisation, metric hierarchy, statistical
analysis plan), a calibrated simulator grounded in real data, and a
black-box parameter optimisation component that searches over button
configuration space to maximise CTR. The project addresses common practical
problems including sample ratio mismatch (SRM), the choice between per-user
and per-impression CTR definitions, and the cost of black-box optimisation.

\section*{Аннотация}
\begin{otherlanguage}{russian}
Проект представляет собой полный пайплайн A/B-тестирования в постановке
«чёрного ящика», применённый к эксперименту с UI-кнопкой. Система
рассматривается как чёрный ящик: входом является назначение пользователю
варианта A (control) или B (treatment), а выходом --- продуктовые метрики
(CTR), рассчитанные по логам событий. Результатом работы являются
формализованная методология эксперимента, откалиброванный симулятор,
основанный на реальных данных, и компонент оптимизации параметров кнопки.
Рассматриваются типовые проблемы: SRM, выбор метрики CTR и стоимость
оптимизации чёрного ящика.
\end{otherlanguage}

% 4. Keywords
\section*{Keywords}
\addcontentsline{toc}{section}{Keywords}
A/B testing; online controlled experiments; black-box simulation; UI button
experiment; CTR; per-impression CTR; SRM; black-box optimisation; SymPy;
calibration

\clearpage

% 5. Introduction
\section{Introduction}

\subsection{Subject Area and Motivation}
Modern digital products improve user experience through stepwise changes to
UI elements, ranking logic, and recommendation features. Online randomised
experiments (A/B tests) are the most reliable decision instrument for
evaluating such changes, because they support causal interpretation of the
observed effect. This project uses a UI button change as the experimental
object, treating the button--user system as a black box to study the
input--output relationship in isolation from the internal mechanism.

\subsection{Problem Statement}
We consider a webpage with a button whose visual parameters (colour, size,
label text) can be varied continuously in $[0, 1]$. A new button configuration
(version B) is proposed. The goal is to determine whether version B improves
the primary product metric (CTR) compared to the current version A, and, given
a calibrated simulator, to identify the configuration that maximises CTR.

\textbf{Statement.} Run an A/B test where users are randomly assigned to A or
B. Using event logs, compute agreed metrics and decide whether to deploy B.
Additionally, query the calibrated black box to find the parameter combination
that maximises expected CTR.

\subsection{Outcomes}
The project delivered:
\begin{itemize}
  \item a formalised description of the experiment, including the unit of
        randomisation, metric hierarchy, and statistical analysis plan;
  \item a Python package \texttt{ab\_blackbox} with symbolic response models
        (SymPy), a black-box simulator, a statistical analysis pipeline, and
        dataset calibration utilities;
  \item a real-data analysis on a publicly available 120\,000-user dataset,
        yielding a statistically significant $+10.0\%$ CTR lift and a
        \textbf{SHIP} decision;
  \item a black-box optimisation component comparing grid search, random
        search, and greedy sweep, with results averaged over 20 seeds.
\end{itemize}

\clearpage

% 6. Literature Review
\section{Literature Review}

The project draws on two bodies of literature: causal inference foundations
and practical online experimentation.

\paragraph{Causal foundations.}
Rubin (1999) establishes the potential outcomes framework and shows that
random assignment is sufficient for causal interpretation of observed
differences~\cite{rubin1999teaching}. This paper provides the theoretical
basis for interpreting the A/B test as a causal experiment rather than a
mere correlation study. Our project adopts Rubin's framework explicitly
in the Assumptions section: SUTVA, ignorability, and positivity are stated
as preconditions, and all three are satisfied by design under user-level
randomisation with a 50/50 split.

\paragraph{Practical online A/B testing.}
Kohavi and Longbotham (2016) describe the end-to-end lifecycle of
online controlled experiments in production systems, including variant
definition, randomisation unit selection, and guardrail
metrics~\cite{kohavi2016online}. Deng et al.\ (2017) extend this with
statistical warnings: naive analysis that ignores the randomisation unit
underestimates variance and overstates significance~\cite{deng2017trustworthy}.
Our project follows both papers by fixing the unit at the user level, using
binary per-user CTR as the primary metric, and running a supplementary
per-impression CTR analysis to confirm that the result is not an artefact
of the metric definition.

\paragraph{Pitfalls and diagnostics.}
Johari et al.\ (2017, 2022) show that continuous monitoring invalidates
classical hypothesis tests and propose always-valid inference
alternatives~\cite{johari2017peeking,johari2022always}. We address this by
using a fixed-horizon stopping rule: no interim analysis is performed, which
keeps the type-I error at the nominal $\alpha = 0.05$. Fabijan et al.\ (2019)
demonstrate that sample ratio mismatch (SRM) is a common and underdiagnosed
data quality problem that can invalidate all downstream
analysis~\cite{fabijan2019srm}. Following their recommendation, our pipeline
runs an SRM chi-square check before any metric analysis.

\clearpage

% ============================================================
\section*{Implementation Overview}
\addcontentsline{toc}{section}{Implementation Overview}
% ============================================================

The practical component is a Python package \texttt{ab\_blackbox} with five
modules and three scripts. Table~\ref{tab:files} lists all files and their
roles.

\begin{table}[ht]
\centering
\caption{Project files and their roles}
\label{tab:files}
\begin{tabular}{lll}
\toprule
\textbf{File} & \textbf{Layer} & \textbf{Role} \\
\midrule
\texttt{ab\_blackbox/model.py}        & Package & Symbolic response models (SymPy) \\
\texttt{ab\_blackbox/simulator.py}    & Package & Black-box oracle \\
\texttt{ab\_blackbox/experiment.py}   & Package & A/B experiment runner \\
\texttt{ab\_blackbox/analysis.py}     & Package & Statistical analysis pipeline \\
\texttt{ab\_blackbox/datasets.py}     & Package & Data loading and calibration \\
\texttt{ab\_blackbox/\_\_init\_\_.py} & Package & Public API \\
\texttt{demo.py}                      & Script  & Synthetic end-to-end demonstration \\
\texttt{main.py}                      & Script  & Real-data pipeline \\
\texttt{optimize.py}                  & Script  & Black-box parameter optimisation \\
\texttt{tests/test\_ab\_blackbox.py}  & Tests   & 14 unit tests \\
\bottomrule
\end{tabular}
\end{table}

\clearpage

% ============================================================
\section{Experimental Design}
% ============================================================

The experiment follows a standard two-variant randomised design. Users are
assigned independently and uniformly at random to control (A) or treatment
(B) at the user level, ensuring consistent exposure throughout the
experiment.

The \emph{experimental object} is a button displayed to users on a webpage.
The treatment modifies three button parameters---colour, size, and label
text---each encoded as a continuous value in $[0, 1]$, where $0$ is the
current default and $1$ is the fully activated alternative. The baseline
experiment sets $\texttt{color} = 1$ while holding $\texttt{size} =
\texttt{text} = 0$, matching the real-world dataset structure where only
one change separates control and test.

The joint system of button and user is treated as a \emph{black box}: the
only observable quantities are the variant assignment and the output event
log. A single call to the black box simulates the full experiment for a
given parameter configuration and returns aggregate metrics. Call count is
tracked to quantify the cost of optimisation.

The dataset used for calibration contains $n = 60{,}000$ users per group,
substantially exceeding the analytically derived minimum of $n = 9{,}458$
required to detect the observed effect at $80\%$ power and $\alpha = 0.05$
(Section~\ref{sec:stat}).

% ============================================================
\section{Metric System}
% ============================================================

Three tiers of metrics are defined prior to data collection.

\paragraph{Primary metric.} Per-user click-through rate (CTR), defined
as the fraction of users who clicked the button at least once:
\[
    \widehat{\mathrm{CTR}}_{\text{user}} = \frac{1}{n} \sum_{i=1}^{n}
    \mathbf{1}[\text{clicks}_i \geq 1].
\]
This is a conservative definition that counts each user at most once,
regardless of how many clicks they made.

\paragraph{Supplementary metric.} Per-impression CTR, defined as the ratio
of total clicks to total page views across all users:
\[
    \widehat{\mathrm{CTR}}_{\text{impression}} =
    \frac{\sum_{i=1}^{n} \text{clicks}_i}{\sum_{i=1}^{n} \text{views}_i}.
\]
This metric uses the full click count and normalises by opportunity (views).
It is reported alongside the primary metric to confirm that the result is
not an artefact of the binary per-user definition. The two metrics are
consistent: per-user lift $+10.0\%$, per-impression lift $+8.3\%$.

\paragraph{Guardrail metric.} The sample ratio $\hat{r} = n_A/(n_A + n_B)$
is monitored before any metric analysis. A significant deviation from
$0.5$ (chi-square test at $\alpha = 0.01$) invalidates the experiment.

% ============================================================
\section{Data Quality and Logging}
% ============================================================

The dataset \texttt{ab\_test\_results\_aggregated\_views\_clicks\_2.csv}
contains $120{,}000$ rows with columns \texttt{user\_id}, \texttt{group},
\texttt{views}, and \texttt{clicks}. Groups are \texttt{control} and
\texttt{test}. The dataset is perfectly balanced ($n_A = n_B = 60{,}000$),
has no missing values, and no user has more clicks than views. No SRM is
present ($p = 1.000$).

The \texttt{views} column records per-user page-view counts (range 1--205,
mean $\approx 5$). It is used to compute the per-impression CTR. The
\texttt{clicks} column records raw click counts per user (range 0--9);
binarised to $\mathbf{1}[\text{clicks} \geq 1]$ for the per-user CTR.

% ============================================================
\section{Statistical Analysis Plan}
\label{sec:stat}
% ============================================================

\paragraph{Primary test.} The treatment effect on per-user CTR is estimated
using a two-proportion $z$-test. Let $p_A$ and $p_B$ denote the true click
probabilities. The null hypothesis is $H_0: p_B = p_A$ against the
two-sided alternative $H_1: p_B \neq p_A$. The test statistic is
\[
    z = \frac{\hat{p}_B - \hat{p}_A}{\sqrt{
        \hat{p}(1-\hat{p})\!\left(\frac{1}{n_A}+\frac{1}{n_B}\right)}},
    \qquad
    \hat{p} = \frac{x_A + x_B}{n_A + n_B},
\]
where $x_A, x_B$ are click counts. The $95\%$ CI on $\Delta = p_B - p_A$
uses the unpooled standard error:
\[
    \Delta \pm z_{0.025}
    \sqrt{\frac{\hat{p}_A(1-\hat{p}_A)}{n_A} +
          \frac{\hat{p}_B(1-\hat{p}_B)}{n_B}}.
\]

\paragraph{Minimum sample size.}
\[
    n = \frac{(z_{\alpha/2} + z_{\beta})^2
        \bigl(p_A(1-p_A) + p_B(1-p_B)\bigr)}{\delta^2},
\]
where $\delta = |p_B - p_A|$. Derived symbolically in \texttt{model.py}
using SymPy and evaluated at the calibrated parameter values.

\paragraph{Bootstrap robustness check.} A non-parametric bootstrap CI
(2000 resamples, vectorised) is computed alongside the $z$-test. Agreement
confirms the $z$-test assumptions are satisfied.

\paragraph{Stopping rule.} The experiment runs for a fixed pre-determined
duration. No interim analysis is performed, which avoids the inflated
type-I error from continuous monitoring~\cite{johari2017peeking}.

% ============================================================
\section{Risks and Limitations}
% ============================================================

\paragraph{Novelty effect.} Users may click the new button simply because
it is unfamiliar. The dataset contains no timestamps, so novelty decay
cannot be measured directly. Monitoring post-deployment CTR would confirm
whether the lift persists.

\paragraph{Assumed parameter weights.} The dataset contains only two
conditions (color=0 vs.\ color=1). The weights for \texttt{size} and
\texttt{text} are not observed and are set equal to the colour effect
($w = 0.0147$) as a \emph{baseline assumption with no empirical basis}.
The optimisation results are therefore illustrative only; follow-up
experiments varying size and text are needed to replace the assumption with
measured values.

\paragraph{Per-user binary CTR.} Using $\mathbf{1}[\text{clicks} \geq 1]$
treats a user with one click identically to a user with nine clicks. The
per-impression metric addresses this, and both tell a consistent story.

\paragraph{Simulation noise.} The simulator adds Gaussian noise to
$p_{\text{click}}$ once per black-box call (noise\_std$= 0.003$ for
optimisation). This models between-experiment variation in population CTR,
not within-experiment confounding. For calibration validation,
noise\_std$= 0$ is used so that the simulated $\Delta$ matches the
calibration target exactly.

% ============================================================
\section{Decision Rules}
% ============================================================

The decision framework has four outcomes:

\begin{enumerate}
    \item \textbf{INVALID}: SRM detected ($p < 0.01$). Halt; investigate.
    \item \textbf{SHIP}: Primary metric significantly positive. Deploy B.
    \item \textbf{REJECT}: Primary metric significantly negative. Do not deploy.
    \item \textbf{INCONCLUSIVE}: Not significant. Increase sample size or duration.
\end{enumerate}

% ============================================================
\section{Assumptions}
% ============================================================

The analysis rests on four assumptions from the potential outcomes
framework~\cite{rubin1999teaching}:

\begin{enumerate}
    \item \textbf{SUTVA}: Each user's outcome depends only on their own
          assignment (no interference between users).
    \item \textbf{Ignorability}: Assignment is independent of potential
          outcomes. Satisfied by design under random assignment.
    \item \textbf{Positivity}: Each user has a non-zero probability of
          assignment to either variant. Satisfied by the 50/50 split.
    \item \textbf{Stable response surface}: Click probability is modelled
          as a deterministic function of button parameters plus i.i.d.\
          noise. Parameter interactions are not modelled.
\end{enumerate}

\clearpage

% ============================================================
\section*{Experimental Results}
\addcontentsline{toc}{section}{Experimental Results}
% ============================================================

Table~\ref{tab:results} reports key metrics for the direct test on real
data and the calibrated simulation (noise\_std$=0$).

\begin{table}[ht]
\centering
\caption{A/B test results: real data and calibrated simulation}
\label{tab:results}
\begin{tabular}{lcccc}
\toprule
\textbf{Source} & \textbf{CTR\,A} & \textbf{CTR\,B}
    & $\boldsymbol{\Delta}$ & \textbf{$p$-value} \\
\midrule
Real data (direct $z$-test)              & 0.1470 & 0.1617 & $+0.0147$ & $<0.0001$ \\
Calibrated simulation (noise\_std$=0$)   & 0.1462 & 0.1623 & $+0.0161$ & $<0.0001$ \\
\bottomrule
\end{tabular}
\end{table}

For the direct test: $z = 7.055$, $95\%$ CI on $\Delta$: $[+0.0106,\,
+0.0188]$, relative lift $+10.0\%$. The per-impression CTR also increased
significantly ($+8.3\%$, $z = 5.52$, $p < 0.0001$), confirming that the
result is not an artefact of the per-user binary definition.

With noise\_std$=0$, the calibrated simulation's $\Delta$ ($+0.0161$)
is close to the real $\Delta$ ($+0.0147$); the small discrepancy is due
to Bernoulli sampling variance with seed 42 at $n = 60{,}000$. The
bootstrap CI $[+0.0121,\, +0.0204]$ excludes zero and agrees directionally
with the $z$-test. No SRM was detected ($p = 1.000$). The decision rule
yields \textbf{SHIP}.

\paragraph{Required sample size.} The minimum per-group $n$ to detect
$\delta = 0.0147$ at $80\%$ power and $\alpha = 0.05$ is $n = 9{,}458$,
derived analytically from the SymPy model. The actual dataset
($n = 60{,}000$) is approximately $6\times$ larger than required.

\paragraph{Black-box optimisation.} Table~\ref{tab:optim} shows results
averaged over 20 independent seeds. All three methods find the same
optimal configuration $(\texttt{color}=1, \texttt{size}=1,
\texttt{text}=1)$; the difference in mean CTR reflects the efficiency
of each search strategy, not a difference in the configuration found.
The theoretical CTR at $(1,1,1)$ from the SymPy model is $0.1912$.

\begin{table}[ht]
\centering
\caption{Optimisation results averaged over 20 seeds. All methods find
config $(1,1,1)$; CTR differences reflect sampling noise, not different
optima. The theoretical CTR at $(1,1,1)$ is $0.1912$ (SymPy).}
\label{tab:optim}
\begin{tabular}{lcccc}
\toprule
\textbf{Method} & \textbf{Calls} & \textbf{Mean CTR}
    & \textbf{Std} & \textbf{Lift vs.\ real test} \\
\midrule
Real experiment (A$\to$B)    & ---  & 0.1617 & ---    & (reference) \\
Grid search                  & 125  & 0.1973 & 0.0038 & $+22.0\%$ \\
Random search (30 calls)     &  30  & 0.1917 & 0.0037 & $+18.5\%$ \\
Greedy 1-D sweep             &  15  & 0.1890 & 0.0061 & $+16.9\%$ \\
SymPy theoretical            &   0  & 0.1912 & ---    & $+18.2\%$ \\
\bottomrule
\end{tabular}
\end{table}

\clearpage

% ============================================================
\section*{Discussion: Interpretation and Effects}
\addcontentsline{toc}{section}{Discussion: Interpretation and Effects}
% ============================================================

\paragraph{Treatment effect.} The test variant produces a statistically
significant increase in per-user CTR of $+0.0147$ (relative lift
$+10.0\%$) and in per-impression CTR of $+0.0025$ ($+8.3\%$). The effect
is robust: confirmed by the parametric $z$-test, the non-parametric
bootstrap, and two complementary metric definitions.

\paragraph{Black-box perspective.} Treating the button--user system as a
black box is methodologically appropriate: no assumptions about the internal
mechanism are required. The causal interpretation relies only on random
assignment~\cite{rubin1999teaching}, which is satisfied by design.

\paragraph{Optimisation vs.\ single test.} The standard A/B test compares
two fixed configurations. The optimisation component extends this to a
search over the three-dimensional parameter space. Averaging over 20 seeds
shows that random search (30 calls) recovers a mean CTR of $0.1917$,
compared to grid search's $0.1973$ (125 calls), illustrating the
efficiency gain from sub-exhaustive methods. The greedy sweep uses only
15 calls but has higher variance ($\sigma = 0.0061$), since it can miss
parameter interactions.

The SymPy theoretical value ($0.1912$) serves as the ground truth: it is
the true model maximum at $(1,1,1)$, computed with zero black-box calls and
zero noise. Any method's mean CTR that exceeds $0.1912$ reflects
noise, not a superior algorithm.

\paragraph{Limitations.} The weights for \texttt{size} and \texttt{text}
are not observed from data and are set to a baseline assumption equal to
the colour weight. Conclusions about the optimal multi-parameter
configuration depend on this assumption. A follow-up experiment would
allow the model to be re-calibrated with observed weights, making the
optimisation results empirically grounded.

\clearpage

% ============================================================
\section*{Conclusion}
\addcontentsline{toc}{section}{Conclusion}
% ============================================================

This project delivered a complete black-box A/B testing pipeline, from
symbolic response modelling (SymPy) through simulation, statistical
analysis, and parameter optimisation. The treatment variant produces a
statistically significant $+10.0\%$ lift in per-user CTR ($+8.3\%$
per-impression), the decision rule yields \textbf{SHIP}, and the
optimisation component---averaged over 20 seeds---identifies configurations
with up to $+22.0\%$ additional lift using 125 calls (grid search) or
$+18.5\%$ with 30 calls (random search).

\clearpage

\printbibliography

\end{document}
