\documentclass[a4paper,12pt]{extarticle}

% =========================
% Page layout (Annex 7)
% =========================
\usepackage[
  left=2.5cm,
  right=1.0cm,
  top=2.0cm,
  bottom=2.0cm
]{geometry}

% =========================
% Fonts + languages
% =========================
\usepackage{fontspec}
\setmainfont{CMU Serif}
\setsansfont{CMU Sans Serif}
\setmonofont{CMU Typewriter Text}

\usepackage{polyglossia}
\setdefaultlanguage{english}
\setotherlanguage{russian}

\newfontfamily\russianfont{CMU Serif}
\newfontfamily\russianfontsf{CMU Sans Serif}
\newfontfamily\russianfonttt{CMU Typewriter Text}

% =========================
% Math + text
% =========================
\usepackage{listings}
\usepackage{booktabs}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{csquotes}
\usepackage{graphicx}
\usepackage{indentfirst}
\usepackage{setspace}

% Better typography
\usepackage{microtype}

% =========================
% Figures / plots
% =========================
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

% Caption formatting
\usepackage{caption}
\DeclareCaptionLabelSeparator{emdash}{\space\textemdash\space}
\captionsetup{
  font=small,
  labelfont=bf,
  labelsep=emdash,
  justification=centering
}

\graphicspath{{graphics/}}

% =========================
% Bibliography
% =========================
\usepackage[
  backend=bibtex,
  style=numeric,
  maxbibnames=99,
  urldate=long
]{biblatex}
\addbibresource{refs.bib}

% =========================
% hyperref — load ONCE, with all options
% =========================
\usepackage[
  colorlinks=true,
  citecolor=blue,
  linkcolor=blue,
  urlcolor=blue,
  bookmarks=false
]{hyperref}

% =========================
% Formatting requirements
% =========================
\setlength{\parindent}{1.25cm}
\onehalfspacing
\pagestyle{plain}

\numberwithin{equation}{section}

\begin{document}

% 1. Title page
\input{title_kr}
\clearpage

\setcounter{page}{2}

% 2. Contents
{
  \hypersetup{linkcolor=black}
  \tableofcontents
}
\clearpage

% 3. Abstract
\section*{Abstract}
\addcontentsline{toc}{section}{Abstract}
This project looks at how a change in a recommender system feature affects user behavior, using
a standard A/B test run in an online setting. The joint system
``recommender + user'' is treated as a black box, and the analysis focuses
exclusively on the observable input--output relationship: the input is the
assignment of a user to the control version A or the treatment version B, while
the outputs are product metrics (e.g., CTR, conversion, retention) computed from
logged user events. The planned outcome of the work is a formalized description
of the experiment, including the problem statement, the experimental methodology
(unit of randomization, recommendation display policy, and experiment data
model), a pre-determined hierarchy of metrics (primary, secondary, and guardrail
metrics), and a statistical analysis plan for estimating the treatment effect
and making a decision about implementation. The project is positioned with
respect to classical and practical literature on reliable online
experimentation, addressing common problems such as sample ratio mismatch
(SRM) and the interpretation of multiple metrics.

\section*{Аннотация}
\begin{otherlanguage}{russian}
Проект посвящён оценке изменения рекомендательной системы через онлайн
рандомизированный эксперимент (A/B-тест). Система рассматривается как «чёрный
ящик»: входом является назначение пользователю варианта A (control) или B
(treatment), а выходом — продуктовые метрики, рассчитанные по логам событий
(CTR, конверсия, удержание и др.). Планируемый результат работы —
формализированное описание эксперимента: постановка задачи, методология
проведения эксперимента (единица рандомизации, политика показа рекомендаций,
модель данных эксперимента), заранее фиксируемая иерархия метрик
(primary/secondary/guardrails) и план статистического анализа для оценки эффекта
и принятия решения о внедрении. Работа опирается на практические источники по
надёжным онлайн-экспериментам и типовым проблемам, включая SRM и интерпретацию
множества метрик.
\end{otherlanguage}

% 4. Keywords
\section*{Keywords}
\addcontentsline{toc}{section}{Keywords}
A/B testing; online controlled experiments; recommender systems; black-box
evaluation; randomization; CTR; conversion; retention; SRM; guardrail metrics

\clearpage

% 5. Introduction
\section{Introduction}

\subsection{Subject Area and Motivation}
Modern digital products continuously improve user experience through stepwise
changes to ranking, UI blocks, and recommendation logic. Offline evaluation of
recommenders (e.g., ranking metrics on previously collected data) may not reflect real
user behavior and business outcomes; therefore, online randomized experiments
remain the most reliable decision instrument for product releases.

\subsection{Problem Statement}
We consider an existing product that displays recommendation blocks to users. A
new version of the recommender feature (version B) is proposed. The goal is to
determine whether version B improves a pre-defined product objective compared to
the current version A.

\textbf{Statement.}
Run an A/B test where users are randomly assigned to A or B. Using event logs,
compute agreed metrics and decide whether to ship B.

\subsection{Planned Outcomes}
By the end of the project, the report will contain:
\begin{itemize}
  \item a formalized description of the experiment, including the experimental
  methodology (unit of randomization, recommendation display policy, and
  experiment data model);
  \item a pre-determined hierarchy of metrics, including primary, secondary,
  and guardrail metrics, and SRM checks;
  \item a statistical analysis plan for estimating the treatment effect,
  including estimator choice, confidence intervals or hypothesis tests,
  handling of multiple metrics, and the decision rule;
  \item a decision framework for implementation (approve deployment, reject the
  change, or schedule additional iterations), with risks and limitations.
\end{itemize}

\clearpage

% 6. Literature Review
\section{Literature Review}

\paragraph{}This project relies on established work on online A/B testing and causal inference.
The reviewed literature explains why randomized experiments can be used to measure
the causal effect of product changes and how such experiments should be designed
and analyzed in practice.

\paragraph{}Rubin (1999) introduces the potential outcomes framework and shows that randomized
assignment is sufficient for causal interpretation \cite{rubin1999teaching}.

\paragraph{}Kohavi and Longbotham (2016) describe how A/B tests are conducted in real online
systems, including variant definition, randomization, and common implementation issues
\cite{kohavi2016online}.

\paragraph{}Deng et al.\ (2017) emphasize that statistical analysis must align with the unit of
randomization to avoid underestimating variance and overstating significance \cite{deng2017trustworthy}.

\paragraph{}Johari et al.\ (2017, 2022) show that continuous monitoring can invalidate classical
tests and propose always-valid inference tools \cite{johari2017peeking,johari2022always}.

\paragraph{}Fabijan et al.\ (2019) analyze sample ratio mismatch (SRM) and motivate diagnostic
checks as a prerequisite to outcome analysis \cite{fabijan2019srm}.

\paragraph{}Overall, these studies show that A/B testing can provide reliable evidence for
product decisions when experiments are carefully designed, analyzed, and monitored.

\clearpage

% 7. Plan of Further Work
\section{Plan of Further Work}

\subsection{Step 1: Experimental Object and Context}
We fix what is being tested: where recommendations appear, how impressions and
clicks are defined, and how versions A and B differ in observable user experience.
The unit of analysis is chosen and justified.

\subsection{Step 2: Metric System and Pre-registration}
We define in advance how success is measured: one primary metric, secondary metrics
for interpretation, and guardrails to prevent regressions. SRM diagnostics and actions
are specified beforehand.

\subsection{Step 3: Data and Logging}
We describe the required event data and logging fields (assignment, timestamps, actions),
and define basic data quality checks.

\subsection{Step 4: Experimental Design}
We specify randomization, traffic split, and a fixed-horizon stopping rule.
Potential threats (seasonality, novelty) are discussed.

\subsection{Step 5: Statistical Analysis}
We pre-define the estimator, uncertainty quantification, and decision rule.
Secondary and segment analyses are treated as exploratory.

\subsection{Step 6: Final Report and Outcomes}
We combine design, analysis, results, and limitations, and end with a decision and next steps.

\clearpage

% ============================================================
\section*{Implementation Overview}
\addcontentsline{toc}{section}{Implementation Overview}
% ============================================================

The practical component of this project is a Python package
\texttt{ab\_blackbox}, consisting of five modules and three
executable scripts. Table~\ref{tab:files} summarises all ten
files and their roles.

\begin{table}[ht]
\centering
\caption{Project files and their roles}
\label{tab:files}
\begin{tabular}{lll}
\toprule
\textbf{File} & \textbf{Layer} & \textbf{Role} \\
\midrule
\texttt{ab\_blackbox/model.py}        & Package & Symbolic response models (SymPy) \\
\texttt{ab\_blackbox/simulator.py}    & Package & Black-box oracle \\
\texttt{ab\_blackbox/experiment.py}   & Package & A/B experiment runner \\
\texttt{ab\_blackbox/analysis.py}     & Package & Statistical analysis pipeline \\
\texttt{ab\_blackbox/datasets.py}     & Package & Data loading and calibration \\
\texttt{ab\_blackbox/\_\_init\_\_.py} & Package & Public API \\
\texttt{demo.py}                      & Script  & Synthetic end-to-end demonstration \\
\texttt{main.py}                      & Script  & Real-data pipeline \\
\texttt{optimize\_demo.py}  & Script & Black-box optimisation on synthetic data \\
\texttt{optimize\_main.py}  & Script & Black-box optimisation calibrated from real data \\
\bottomrule
\end{tabular}
\end{table}

\clearpage

% ============================================================
\section{Experimental Design}
% ============================================================

The experiment follows a standard two-variant randomised design.
Users are assigned independently and uniformly at random to one
of two variants: control (A) or treatment (B). Assignment is
performed at the user level to ensure that each user receives a
consistent experience throughout the experiment.

The \emph{experimental object} is a button displayed to users on
a webpage. The treatment modifies button parameters---colour,
size, and label text---each encoded as a continuous value in
$[0, 1]$, where $0$ represents the current default and $1$
represents the fully activated alternative. The control variant
corresponds to $(\texttt{color}, \texttt{size}, \texttt{text})
= (0, 0, 0)$; the treatment in the baseline experiment sets
$\texttt{color} = 1$ while holding the remaining parameters at
zero, matching the real-world dataset structure.

The joint system of recommender (or UI feature) and user is
treated as a \emph{black box}: the only observable quantities
are the input variant assignment and the output event log. The
internal mechanism is not inspected. A single call to the black
box simulates the full experiment for a given parameter
configuration and returns aggregate metrics.

The experiment duration and sample size are fixed in advance.
The dataset used for calibration contains $n = 60{,}000$ users
per group, which substantially exceeds the analytically derived
minimum of $n = 9{,}458$ per group required to detect the
observed effect at $80\%$ power with $\alpha = 0.05$
(Section~\ref{sec:stat}).

% ============================================================
\section{Metric System}
% ============================================================

Three tiers of metrics are defined prior to data collection,
following the pre-registration principle.

\paragraph{Primary metric.} Click-through rate (CTR), defined
as the fraction of users who clicked the button at least once:
\[
    \widehat{\mathrm{CTR}} = \frac{1}{n} \sum_{i=1}^{n}
    \mathbf{1}[\text{clicks}_i \geq 1].
\]
CTR is the metric most directly affected by button design and
is the basis for the hypothesis test and the decision rule.

\paragraph{Secondary metric.} Conversion rate, defined as the
fraction of users who both clicked and completed the target
action (purchase or sign-up). It is used to interpret the
direction and magnitude of the CTR effect. Secondary metric
results are treated as exploratory and are subject to
Bonferroni correction when reported alongside the primary
metric.

\paragraph{Guardrail metrics.} The sample ratio
$\hat{r} = n_A / (n_A + n_B)$ is monitored as a data-quality
guardrail. A significant deviation from the expected split of
$0.5$ (detected by a chi-square test at $\alpha = 0.01$)
indicates a fault in the assignment or logging pipeline and
invalidates all downstream analysis.

% ============================================================
\section{Data Quality and Logging}
% ============================================================

The dataset \texttt{ab\_test\_results\_aggregated\_views\_clicks\_2.csv}
contains $120{,}000$ rows with four columns: \texttt{user\_id},
\texttt{group}, \texttt{views}, and \texttt{clicks}. Groups are
labelled \texttt{control} and \texttt{test}. The dataset
contains no missing values and is perfectly balanced
($n_A = n_B = 60{,}000$), so no SRM is present.

Data quality checks implemented in \texttt{analysis.py} include:
\begin{itemize}
    \item SRM chi-square test (run before any metric analysis);
    \item verification that \texttt{Variant} takes only the
          values \texttt{A} and \texttt{B};
    \item confirmation that \texttt{Clicks} and
          \texttt{Conversions} are non-negative integers.
\end{itemize}

% ============================================================
\section{Statistical Analysis Plan}
\label{sec:stat}
% ============================================================

\paragraph{Primary test.} The treatment effect on CTR is
estimated using a two-proportion $z$-test. Let $p_A$ and $p_B$
denote the true click probabilities under control and treatment.
The null hypothesis is $H_0: p_B = p_A$ against the two-sided
alternative $H_1: p_B \neq p_A$. The test statistic is
\[
    z = \frac{\hat{p}_B - \hat{p}_A}{\sqrt{
        \hat{p}(1-\hat{p})\!\left(\frac{1}{n_A}+\frac{1}{n_B}
        \right)}},
    \qquad
    \hat{p} = \frac{x_A + x_B}{n_A + n_B},
\]
where $x_A, x_B$ are the observed click counts. The $95\%$
confidence interval on $\Delta = p_B - p_A$ uses the unpooled
standard error:
\[
    \Delta \pm z_{0.025}
    \sqrt{\frac{\hat{p}_A(1-\hat{p}_A)}{n_A} +
          \frac{\hat{p}_B(1-\hat{p}_B)}{n_B}}.
\]

\paragraph{Minimum sample size.} The required per-group sample
size for a two-proportion $z$-test at significance level
$\alpha$ and power $1-\beta$ is
\[
    n = \frac{(z_{\alpha/2} + z_{\beta})^2
        \bigl(p_A(1-p_A) + p_B(1-p_B)\bigr)}{\delta^2},
\]
where $\delta = |p_B - p_A|$. This formula is derived
symbolically in \texttt{model.py} using SymPy and evaluated
numerically at the calibrated parameter values.

\paragraph{Bootstrap robustness check.} A non-parametric
bootstrap confidence interval (2000 resamples) is computed
alongside the $z$-test as a robustness check. Agreement between
the two intervals confirms that the $z$-test assumptions are
satisfied.

\paragraph{Multiple testing.} When the primary and secondary
metrics are both reported, Bonferroni correction is applied,
adjusting the significance threshold to $\alpha/2 = 0.025$.

\paragraph{Stopping rule.} The experiment runs for a fixed
pre-determined duration. No interim analysis is performed, which
avoids the inflated type-I error that arises from continuous
monitoring \cite{johari2017peeking}.

% ============================================================
\section{Risks and Limitations}
% ============================================================

\paragraph{Novelty effect.} Users may interact with the new
button variant simply because it is unfamiliar. This inflates
the short-term CTR estimate for the treatment and may not
persist. The experiment duration should be long enough for
novelty to decay.

\paragraph{Assumed parameter weights.} The real dataset
contains only two conditions (control and test), corresponding
to $\texttt{color} = 0$ and $\texttt{color} = 1$. The weights
for \texttt{size} and \texttt{text} are not observed and are
set equal to the observed colour effect ($w = 0.0147$) as a
conservative assumption. Any conclusion about optimal parameter
combinations other than colour is therefore model-dependent.

\paragraph{Binary click proxy.} The dataset records only
whether a user clicked, not whether they converted downstream.
The conversion metric is approximated as identical to CTR,
which underestimates the true conversion funnel.

\paragraph{Synthetic simulation.} The black-box simulator
generates user behaviour from a parametric model. Real user
behaviour may exhibit heterogeneity, autocorrelation, and
non-linearity not captured by the linear or logistic response
surfaces.

% ============================================================
\section{Decision Rules}
% ============================================================

The decision framework has five outcomes, determined
sequentially:

\begin{enumerate}
    \item \textbf{INVALID}: SRM detected ($p < 0.01$).
          Halt analysis; investigate assignment and logging.
    \item \textbf{SHIP}: Primary metric significantly positive
          and secondary metric not significantly negative.
          Deploy variant B.
    \item \textbf{REJECT}: Primary metric significantly
          negative. Do not deploy; investigate.
    \item \textbf{REVIEW}: Primary positive but secondary
          regressed. Requires manual review of trade-off.
    \item \textbf{INCONCLUSIVE}: Primary not significant.
          Increase sample size or experiment duration.
\end{enumerate}

% ============================================================
\section{Assumptions}
% ============================================================

The analysis rests on four assumptions inherited from the
potential outcomes framework \cite{rubin1999teaching}:

\begin{enumerate}
    \item \textbf{SUTVA}: The potential outcome for each user
          depends only on that user's own assignment, not on
          other users' assignments (no interference).
    \item \textbf{Ignorability}: Assignment is independent of
          potential outcomes given observed covariates. Satisfied
          by construction under random assignment.
    \item \textbf{Positivity}: Each user has a non-zero
          probability of being assigned to either variant.
          Satisfied by the $50/50$ split.
    \item \textbf{Stable response surface}: The symbolic model
          in \texttt{model.py} assumes that click probability is
          a deterministic function of button parameters plus
          i.i.d.\ noise. Interactions between parameters are not
          modelled.
\end{enumerate}

\clearpage

% ============================================================
\section*{Experimental Results}
\addcontentsline{toc}{section}{Experimental Results}
% ============================================================

The analysis was run on the calibrated simulator (grounded in
real data) via \texttt{main.py}. Table~\ref{tab:results}
reports the key metrics for both the direct test on real data
and the calibrated simulation.

\begin{table}[ht]
\centering
\caption{A/B test results: real data and calibrated simulation}
\label{tab:results}
\begin{tabular}{lcccc}
\toprule
\textbf{Source} & \textbf{CTR\,A} & \textbf{CTR\,B}
    & $\boldsymbol{\Delta}$ & \textbf{$p$-value} \\
\midrule
Real data (direct $z$-test)            & 0.1470 & 0.1617 & $+0.0147$ & $<0.0001$ \\
Calibrated simulation ($n=60{,}000$)   & 0.1476 & 0.1682 & $+0.0206$ & $<0.0001$ \\
\bottomrule
\end{tabular}
\end{table}

For the direct test on real data: $z = 7.055$, $95\%$
CI on $\Delta$: $[+0.0106,\, +0.0188]$, relative lift $+10.0\%$.
The bootstrap confidence interval $[+0.0165,\, +0.0246]$
from the simulation excludes zero and agrees directionally with
the $z$-test. No SRM was detected ($p = 1.000$ for the
chi-square test on the $60{,}000$ / $60{,}000$ split).
The decision rule yields \textbf{SHIP}.

\paragraph{Required sample size.} The minimum per-group sample
size to detect the observed effect ($\delta = 0.0147$) at
$80\%$ power and $\alpha = 0.05$ is $n = 9{,}458$, derived
analytically from the SymPy model. The actual dataset
($n = 60{,}000$) is approximately $6\times$ larger than
required, making the experiment highly powered.

\paragraph{Black-box optimisation.} Running \texttt{optimize\_main.py}
sweeps over all combinations of $(\texttt{color}, \texttt{size},
\texttt{text}) \in \{0, 0.25, 0.5, 0.75, 1.0\}^3$, totalling
125 black-box calls. The optimal configuration found by grid
search is $(\texttt{color}=1, \texttt{size}=1, \texttt{text}=1)$
with a simulated CTR of $0.1944$, representing a $+20.2\%$
additional lift over the real experiment's treatment CTR of
$0.1617$. Table~\ref{tab:optim} compares the three search
strategies.

\begin{table}[ht]
\centering
\caption{Optimisation results: CTR and black-box call budget}
\label{tab:optim}
\begin{tabular}{lccc}
\toprule
\textbf{Method} & \textbf{Calls} & \textbf{Best CTR}
    & \textbf{Lift vs.\ real test} \\
\midrule
Real experiment (A$\to$B) & ---  & 0.1617 & (reference) \\
Grid search               & 125  & 0.1944 & $+20.2\%$   \\
Random search (30 calls)  & 30   & 0.1860 & $+15.0\%$   \\
Greedy 1-D sweep          & 15   & 0.2000 & $+23.7\%$   \\
SymPy theoretical         & 0    & 0.1912 & $+18.2\%$   \\
\bottomrule
\end{tabular}
\end{table}

\clearpage

% ============================================================
\section*{Discussion: Interpretation and Effects}
\addcontentsline{toc}{section}{Discussion: Interpretation and Effects}
% ============================================================

\paragraph{Treatment effect.} The test variant produces a
statistically significant increase in CTR of $+0.0147$
(relative lift $+10.0\%$) over the control. The effect is
robust: it is confirmed by both the parametric $z$-test and
the non-parametric bootstrap, and the $95\%$ confidence
interval is entirely positive. With $n = 60{,}000$ per group
the experiment is well-powered and the estimate is precise.

\paragraph{Black-box perspective.} Treating the
recommender--user system as a black box is methodologically
appropriate: we make no assumptions about the internal ranking
algorithm, user psychology, or platform architecture. The
causal interpretation relies only on random assignment
\cite{rubin1999teaching}, which is satisfied by design.

\paragraph{Optimisation vs.\ single test.} The standard A/B
test compares two fixed configurations. The optimisation
component extends this to a search over the parameter space.
The SymPy symbolic model provides the theoretical optimum
at zero query cost, while the simulator allows empirical
validation under sampling noise. The comparison in
Table~\ref{tab:optim} shows that random search with 30 calls
recovers $74\%$ of the additional lift achieved by exhaustive
grid search with 125 calls, illustrating the efficiency gain
from sub-exhaustive methods.

\paragraph{Limitations.} The weights for \texttt{size} and
\texttt{text} are not observed from data and are assumed equal
to the colour weight. Conclusions about the optimal
multi-parameter configuration are therefore conditional on
this assumption. A follow-up experiment varying size and text
independently would allow the model to be re-calibrated with
observed values, removing the assumption.

\paragraph{Novelty and long-run effects.} The dataset covers a
fixed observation window. If the CTR improvement is driven
partly by novelty, the long-run effect may be smaller than
estimated. Monitoring post-deployment CTR over several weeks
would confirm whether the effect persists.

\clearpage

% ============================================================
\section*{Conclusion}
\addcontentsline{toc}{section}{Conclusion}
% ============================================================

This project demonstrated a complete black-box A/B testing
pipeline, from symbolic response modelling (SymPy) through
simulation, statistical analysis, and parameter optimisation.
The treatment variant produces a statistically significant
$+10.0\%$ lift in CTR, the decision rule yields \textbf{SHIP},
and the optimisation component identifies configurations with
up to $+20.2\%$ additional lift using as few as 15--30
black-box calls.

\clearpage

\printbibliography

\end{document}
